---
title: "TUMI Microbiome/Data Science Weekly Meeting - Introduction to ggplot2"
author: "Maureen Carey"
date: "6/1/2021"
output: html_document
---

This tutorial is an introduction to make figures programmatically in R with ggplot2 and uses data from Carey et al.^[https://github.com/maureencarey/metabolomics]

# Prep computational environment

Set path. This will change on your computer - indicate the folder where you saved the data.

```{r setup, include = F}
knitr::opts_chunk$set(echo = TRUE)

path = "/Users/mac9jc/Documents/Documents - somH12CXJ8GJV40/work/projects/other_tutorials/ggplot2/"

knitr::opts_knit$set(root.dir = path)

```

```{r, eval = F}
knitr::opts_chunk$set(echo = TRUE)

path = "/path/to/data/"

knitr::opts_knit$set(root.dir = path)

```

Load packages.

```{r,warning=TRUE, message=FALSE}

library(tidyverse) #dplyr, tidyr
load("for_ggplot2_tutorial.RData")

```

Let's say you want to make a publication-quality figure like this:

```{r, echo = F}

p = ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~group_use, scales = "free") +
  coord_flip() + 
  theme_minimal() +
  theme(axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "bottom") +
  scale_color_manual(values = c("black","red")) +
  scale_fill_manual(values = c("white","grey")) +
  scale_y_log10() + 
  ylab('abundance (log10)') + xlab('metabolites') +
  labs(fill = "Genetic Background:", color = "Drug Treatment:")
p

```

-------

The biggest strength of making figures programmatically (as opposed to Excel or Prism, or other approaches that don't use code) is that your figures are 100% reproducible and customizable and all errors are traceable. Making figures in Excel or Prism may be easier to start, but there is no documentation (i.e. code) of what you did.

`ggplot2`, the package used here, is just one way to do it and my personal favorite. Strengths of `ggplot2` specifically are that its relatively easy to make publication quality figures and many application-focused packages provide even-easier-to-use wrappers for ggplot plotting. 

My favorite feature of `ggplot2` is how you can layer on new information or plotting approaches. For example, in the plot above, we first specified what data would be used and the x- and y-axes here:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background))

```

Next, we told the program what type of plot to make (boxplot) so it plotted the data (specified in the first step) using this approach:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot() 

```

We can also hide all outliers on the plot with this:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA)

```

However, we are actually interested in separating each type of metabolite into groups, like this:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~group_use, scales = "free") 

```

But these colors are rather uninterpretable, so let's change them:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~group_use, scales = "free") + 
  scale_color_manual(values = c("black","red")) +
  scale_fill_manual(values = c("white","grey")) 

```

I prefer the plot to be flipped, so we can do that here:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~group_use, scales = "free") +
  scale_color_manual(values = c("black","red")) +
  scale_fill_manual(values = c("white","grey")) +
  coord_flip()

```

Let's now remove the metabolite identifiers since they are not terribly useful:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~group_use, scales = "free") + 
  scale_color_manual(values = c("black","red")) +
  scale_fill_manual(values = c("white","grey")) +
  coord_flip() + 
  theme(axis.ticks.x = element_blank(),
        axis.text.y = element_blank())

```

Next, let's change the scale of the 'abundance' axis to make it clearer:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA) + 
  facet_wrap(~group_use, scales = "free") + 
  scale_color_manual(values = c("black","red")) +
  scale_fill_manual(values = c("white","grey")) +
  coord_flip() + 
  theme(axis.ticks.x = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_log10()

```

Let's now change the labels for the x- and y-axes:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA) + 
  facet_wrap(~group_use, scales = "free") + 
  scale_color_manual(values = c("black","red")) +
  scale_fill_manual(values = c("white","grey")) +
  coord_flip() + 
  theme(axis.ticks.x = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_log10() + 
  ylab('abundance (log10)') + xlab('metabolites')

```

And change the location and capitalization on the legend:

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA) + 
  facet_wrap(~group_use, scales = "free") + 
  scale_color_manual(values = c("black","red")) +
  scale_fill_manual(values = c("white","grey")) +
  coord_flip() + 
  theme(axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "bottom") +
  scale_y_log10() + 
  ylab('abundance (log10)') + xlab('metabolites') +
  labs(fill = "Genetic Background:", color = "Drug Treatment:")

```

And lastly, let's get rid of the grey background.

```{r}

ggplot(data = plot_these, 
           aes(x = metabolite, y = abundance, color = drug_treatment, fill = genetic_background)) +
  geom_boxplot(outlier.shape = NA) + 
  facet_wrap(~group_use, scales = "free") + 
  scale_color_manual(values = c("black","red")) +
  scale_fill_manual(values = c("white","grey")) +
  coord_flip() +
  theme_minimal() +
  theme(axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "bottom") +
  scale_y_log10() + 
  ylab('abundance (log10)') + xlab('metabolites') +
  labs(fill = "Genetic Background:", color = "Drug Treatment:") 

```

(Notice how we generally added each layer to the end of the code except for 'theme_minimal'. That is because 'theme_minimal' will override our custom 'theme' layer where we hid the ticks and text on one axis.)

----------

Ok, let's try one more! Here is the final plot:

```{r, echo = F}

p = ggplot(new_plot_these) + 
  geom_point(aes(x = minus_PUR, y = plus_PUR), color = "blue") +
  geom_point(aes(x = minus_BAT, y = plus_BAT), color = "red") +
  ylab('metabolite abundance with drug treatment') + 
  xlab('metabolite abundance withOUT drug treatment') +
  scale_x_log10() + scale_y_log10() +
  annotate(geom = "text", label = "genetic background:\ndrug resistant in red\ndrug sensitive in blue", x =.8, y = 4) +
  annotate(geom = "segment", x = 0.5, xend = 5, y = 0.5, yend = 5,
           colour = "grey", linetype = "dashed")
p

```

Layer 1: Data

```{r}

ggplot(new_plot_these)

```

Layer 2: Data & type of plot.

Notice here we put the x and y variables into the type of plot layer; this is because our x and y variables are different for the two sets of points!

```{r}

ggplot(new_plot_these) + 
  geom_point(aes(x = minus_PUR, y = plus_PUR), color = "blue") +
  geom_point(aes(x = minus_BAT, y = plus_BAT), color = "red")

```

Layer 3: Rename axes. 

```{r}

ggplot(new_plot_these) + 
  geom_point(aes(x = minus_PUR, y = plus_PUR), color = "blue") +
  geom_point(aes(x = minus_BAT, y = plus_BAT), color = "red") +
  ylab('metabolite abundance with drug treatment') + 
  xlab('metabolite abundance withOUT drug treatment') 

```

Layer 4: Scale axes. 

```{r}

ggplot(new_plot_these) + 
  geom_point(aes(x = minus_PUR, y = plus_PUR), color = "blue") +
  geom_point(aes(x = minus_BAT, y = plus_BAT), color = "red") +
  ylab('metabolite abundance with drug treatment') + 
  xlab('metabolite abundance withOUT drug treatment') +
  scale_x_log10() + scale_y_log10() 

```

Layer 5: Add custom legend for colors.

```{r}

ggplot(new_plot_these) + 
  geom_point(aes(x = minus_PUR, y = plus_PUR), color = "blue") +
  geom_point(aes(x = minus_BAT, y = plus_BAT), color = "red") +
  ylab('metabolite abundance with drug treatment') + 
  xlab('metabolite abundance withOUT drug treatment') +
  scale_x_log10() + scale_y_log10() +
  annotate(geom = "text", label = "genetic background:\ndrug resistant in red\ndrug sensitive in blue", x =.8, y = 4) 

```

Layer 6: Add dotted line to help with interpretation.

```{r}

ggplot(new_plot_these) + 
  geom_point(aes(x = minus_PUR, y = plus_PUR), color = "blue") +
  geom_point(aes(x = minus_BAT, y = plus_BAT), color = "red") +
  ylab('metabolite abundance with drug treatment') + 
  xlab('metabolite abundance withOUT drug treatment') +
  scale_x_log10() + scale_y_log10() +
  annotate(geom = "text", label = "genetic background:\ndrug resistant in red\ndrug sensitive in blue", x =.8, y = 4) +
  annotate(geom = "segment", x = 0.5, xend = 5, y = 0.5, yend = 5,
           colour = "grey", linetype = "dashed")

```
